Research paper

### 2013-01-05 Research paper.tex
Revision on the basis of Maartens comments

When one of Starfleet's vessels meets an alien ship in areas where no man has gone before, they can (almost) always communicate with these aliens (almost) instantly; they speak to them in English, and these aliens answer in English. This is possible because, according to the Star Trek lore, the protagonists make use of a so-called Universal Translator. This Universal Translator can (1) recognize speech in every language, (2) translate what it has 'heard' to any other language, (3) make its results audible with speech synthesis and (4) learn itself new languages when needed. Unfortunately, like many technologies in Star Trek, this is far from reality: there is lots of active research at the first three tasks in particular (and even some of the fourth, see \citet{biemann11}), but no software currently in existence can do this as perfect as the Universal Translator. \\\indent
In this paper, I will give a detailed overview of the research project Colibri, which basically is an attempt to improve the technology needed for the second task: automatically translating text from one language to another. More specifically, the project investigates the effects of various translation units; for example, whether making use of constructions with one or more gaps or context improves translation quality. Importantly, the translation are not done on the basis of explicit 'human' knowledge about language and grammar ('linguistic theory'), but are distilled from large amounts of text ('text corpora'). The Colibri project is carried out by Maarten van Gompel and supervised by Antal van den Bosch.
\\\indent
In section 2, I'll give a short overview of what has been achieved in machine translation, and in section 3 the current project will be described. This order was chosen to show that Colibri builds on previous research directly. I give more information on the methodology in section 4, and describe the current and future station section 5. Section 6 elaborates on Colibri's relevance, both for science and society, and section 7 describes some strengths and weaknesses of the project. The paper is concluded in section 8.

@@

The horizontal arrows represent the approaches. I'll go through them from bottom to top:

\begin{itemize}
\item \emph{The direct approach} does very little source language analysis, and could thus be viewed as the automatical version of looking up every word in the source language in a dictionary and replacing it with the corresponding word in the target language. This often doesn't yield very good results: if I for example would have translated this paper from my mother tongue Dutch to English word for word, 'would we now not real a good result have'. It should be noted, however, that the direct approach mostly is more complicated than this; for example, in some cases morphological analysis is needed before it can be looked up in a dictionary, and more morphological analysis is needed for the translation word to fit in the sentence. Word reordering rules can also be included.
\item \emph{The syntactic approach} tries to find out the syntactic structure of the source sentence before it starts translating, using a parser. This structure is then translated to the target language, and filled with translated words.
\item \emph{The semantic approach} goes even one step further and also tries to grasp the meaning of a sentence, roughly. A sentence like \emph{Mummy kisses daddy} for example entails that mummy is the agent, and daddy the patient of the sentence. After having identified these semantic roles, the algorithm would then look how this meaning is expressed in the target language, find the approriate syntactic structure for that, and fill it with the translated words.
\item \emph{The interlingua approach}. All of the approaches so far have in common that they require a seperate rule set for each language pair; no or very few rules included for translations from English to German can also be used for translations from English to French. This means that for every time a new language is added to the system, the amount of rules that have to be added grows exponentially. The Interlingua approach is an attempt to solve this problem, by pairing up every language with a formal representation of meaning, the Interlingua. This means a translation from English to German would entail a translation from English to the Interlingua (using the ruleset English $\rightarrow$ Interlingua) and then a translation from the Interlingua to German (using the ruleset Interlingua $\rightarrow$ German). When translating from English to French, the same two steps would be taken, and the same ruleset English $\rightarrow$ Interlingua can be used.
\end{itemize}

I should add that translation system focussing on one such approach are rare. As \citet{jm09} put it: 'Real systems tend to involve combinations of elements from these three architectures; thus, each is best thought of as a point in an algorithmic design space rather than as an actual algorithm.' The Systran system \citep{hs92,senellartea01}, for example combines the first three approaches.\\\indent
An advantage of the rule-based approach is that one has a lot of control over the output. In theory, if something has gone wrong, all one has to do is find out which rule or dictionary item caused the problem and improve it. In practice, however, it is almost impossible to have a dictionary and rule system good enough for more complicated sentences, and things like rule interactions and ambiguity make everything even more complicated. The next two approaches therefore replace hand-crafted rules with automatically generated ones.

@@

Unlike the ruled-based approach, the statistical approach creates its own 'knowledge', using bilingual text corpora. With this knowledge, the algorithm is able to generate various translation hypotheses, from which the best one is chosen with the help of various statistical measures. I will explain these stages in more detail:

\begin{itemize}
\item \emph{Corpus alignment.} Only having a bilingual corpus is not enough: for such a corpus to be useful one also needs to know which parts of the sentence in language A correspond to which parts of the sentence in language B. This is not only complicated by different word orders, but also by concepts that in language A take more or less words than in language B. After aligning words, it is possible to form phrases and align these on the basis of the word-alignment info. \citet{koehn03} show this provides better results than word-based translation.

\item \emph{Hypothesis generation.} We now have some sort of automatically generated dictionary: for each word in language A, we know which word in language B is often used for it. On the basis of this knowledge, we can generate various possible translations. How many and which hypotheses are generated differs per approach; you could for example limit the generated hypotheses on the basis of the evaluation of previously generated hypotheses.

@@

\item \emph{The language model.} The language model predicts to what extent the generated sentence is expected for the target language; in other words, is this sentence a well-formed sentence or not? Because the unlimited number of ways words can be combined into sentences, it is unlikely the sentence under investigation will actually be found in a corpus, so the language model tries to do this prediction on the basis of how often smaller parts of the sentence (n-grams) occur in a corpus.

@@

The algorithm that performs the last two steps, hypothesis generation and evaluation, is called a \emph{decoder}.

@@

For my overview of the example-based approach, I'll focus on the memory-based version of the example-based approach in particular, because that is the version the Colibri project uses. Other versions of example-based machine translation might vary slightly. \\\indent
The memory-based approach is in many ways similar to the statistical approach - it could even easily be argued to be a part of it. For example, the memory based approach also generates translations on the basis of large bilingual corpora, and also works with a decoder to mold the various translation pieces found into a sentence. The main difference, however, is that it is context-sensitive; the idea is that the system will find another translation for \emph{left} in the sentence \emph{The toilets are at your left hand} than for \emph{left} in the sentence \emph{Elvis has left the building} simply because they occur in different contexts. \\\indent
In the memory-based approach, this context-sensitivy is achieved with the help of machine learning: by showing a so-called classifier lots of examples of something, and telling it to which class it belongs, the classifier is able to predict the class of new examples. For example, we can give a classifier the external characteristics of 10.000 humans (length, type of cloths, length of hair, etc.), and also tell it whether these humans are male or female. If we then ask for the sex of a relatively short human, wearing a dress, with long hair, the classifier will probably be able to tell us this is a woman, because most of the examples with similar features also were women. It thus is frequency that plays a central role here.\\\indent
While this example only has two classes, it is in principle possible to use an unlimited number of classes. This is exactly what the memory-based approach to machine translation does: on the basis of a particular word and its context (for example, its direct neighbours), a classifier in machine translation returns as a class the most likely n-gram. Thus, instead of on human characteristics the classifiers are trained on words and their context from the source language, and instead of sex it uses n-grams from the target language as classes. \\\indent
There are two types of classifaction methods: \emph{eager learning} and \emph{lazy learning}. Whereas eager learning algorithms try to create a small, abstract, general model, filtering out infrequent cases, lazy learning algorithms take into account everything they encounter. According to \citet{dvdb05}, for natural language processing the lazy learning approach works best, because infrequent cases are an important part of the knowledge needed. \\\indent
Like I did with statistical machine translation, I will now give an overview of the memory-based machine translation process, adapted from \citet{vdbb09}. For this example, I will describe a translator that uses trigrams, but all kinds of n-grams could be used here, even n-grams of variable length. In fact, \citet{vangompel09}, \citet{vangompelea09} and \citet{vangompel11} showed that better results can be achieved using phrases of variable length; investigating the nature of multi-word units in translation therefore is one of the main goals of the Colibri project.

\begin{itemize}
\item \emph{Training.} Training is done only once and \emph{before} actual translation, and could therefore be seen as 'preparing' the translation software for its task. It consists of (1) aligning the bilingual corpus, using the word alignment algorithms described in the previous sections, (2) decomposing the complete corpus into trigrams, and (3) creating a language model on the basis of this. The trigrams of the source language constitute the examples, the trigrams of the target language are the classes.

\item \emph{Local classification.} Classification means (1) decomposing the source text into trigrams and (2) classifying each trigram, based on the language model created in the training phase. The output are the classes, i.e. the trigrams, the classifiers associates with each source trigram.

@@

@@

The gap here is filled with 'his', but the construction 'from \_\_ point of view' can also be filled with many other words ('my', 'the', 'another', 'Obama's', etc.). This shows the advantage of skipgrams over n-grams for machine translation: whereas n-grams would only have been able to capture the construction 'point of view', because the word directly before 'point of view' is variable, skipgrams are also able to recognize that the word 'for' is part of the construction. Frequent examples of the skipgrams (I expect the variant with 'my' to be very frequent, for instance) of course \emph{could} be recognized by the n-gram approach.\\\indent
Of course, not every possible group of consecutive words is a construction; what makes some groups special? The exact nature of linguistic constructions has been the subject of many linguistic publications and even an entire group of linguistic frameworks (construction grammar, see \citet{goldberg95} for a well-known example). For this project, however, it suffices to say that constructions emerge because some combinations of words are more frequent than others. As we've seen in the previous section, taking into account frequency in the training material is how classifiers are able to do their job.

@@

\subsubsection{Identifying and extracting constructions in a language}
When using large corpora, the process of discovering simple n-grams might already take quite a lot of time; adding variability in lenght and one or multiple gaps to the n-grams makes the process even more - in fact, many times more - complex. To reduce the processing load, two techniques are used:

\begin{itemize}
\item Optimization of how the various the n-grams and skipgrams are saved in memory. By reserving the labels that don't take a lot of memory for the most frequent grams, a lot of memory can be saved.
\item For trigrams and larger, taking into account the frequencies of the parts. If these parts do not exist or are very infrequent, there is no use in looking for the whole - if even the parts don't exist, the whole won't exist for sure and doesnt need to be investigated any further.
\end{itemize}

\subsubsection{Identifying aligned constructions between language-pairs}
Once the constructions in both parts of a bilingual corpus are identified, these constructions need to be linked. Colibri aligns constructions by aligning the individual words first and constructions later, and uses a separate algorithm to align skipgrams.\\\indent
It should be noted that this approach is the result of various investigations. For example, instead of the classical approach of aligning constructions by aligning the individual words first, as described by \citep{koehn03}, it was attempted to try to align the constructions directly. To achieve this, one can for example simply measure how often the constructions found in both languages occur together (that is, in both the source and the target language). Because at this stage the decoder was not yet ready (see subsection 4), the decoder from translation system MOSES \citep{koehnea07} was used to evaluate the results. These results showed that the approach using constructions for alignment could not provide results as good as the approach using words; therefore, it was decided to go back to the classical method. It should be noted that, because the MOSES decoder was used, the influence of skipgrams on alignment could not be measured. However, because they increase complexity, they are not expected to improve the alignments.\\\indent
When support for skipgrams was added to the implementation of the classical approach, it turned out that including skipgrams in the alignment process had negative effects on the results. To fix this, a algorithm specialized for aligning skipgrams was developed, which indeed was able to find good translations.

\subsubsection{Memory-based classification using construction experts}
As may have become clear from the previous sections, the Colibri project uses machine learning techniques to get from the source language construction to the target language constructions. More specifically, Colibri uses little expert classifiers, trained for only one construction, inspired by the effectivity of this approach in word sense ambiguation \citep{vangompel10}. The results of this approach are compared to the results of a combined classifier approach and statistical machine translation.

\subsubsection{Reassembling constructions into a coherent target sentence}
A decoder was developed from scratch, following state-of-the-art techniques as described by \citet{koehn03}. Support for skipgrams, of course not available in standard decoders, was added.

@@

Because of Colibri's practical nature, its methodology forms the core of the project and has been explained in detail in the previous sections, making this section largely superfluous. However, this might be the right place to note that Colibri's unusual methodology also makes its time planning quite different from more conventional research projects: whereas other projects would go through stages like (1) reading the previous literature, (2) building a theoretical model, (3) testing the model, for example with experiments and (4) refining the model on the basis of the results, Colibri takes an entirely different approach. The first year has almost completely been devoted to implementing ideas, resulting in a fully functional machine translation system. With this technology in place, the years after that can be used refining (and possibly extending) this technology, as well as evaluating its results.

@@

At the moment, the functional machine translation system described in section \ref{builder} is mostly finished. It is mainly based on the open source translation system MOSES, and indeed manages to produce similar results. However, as was explained in section \ref{Description}, Colibri is an attempt to improve the quality of machine translation by adding skipgrams and memory-based learning. Unfortunately it has turned out that both do not add much to the translation quality: memory-based learning only improves the results a little bit, and the skipgrams are mostly ignored by the decoder. Apparently, translation hypotheses without skipgrams are more likely to be good translations in most of the cases. \\\indent
This means Colibri's goal, improving translation quality, has not yet been achieved. It is possible that larger improvements in translation quality \emph{can} be achieved by changing things like corpora or even languages used, but if this is not the case, new possibilities for improvements have to be investigated. One such possibility is to include a system that finds the correct translations of words on the basis of particular keywords in a sentences. This system should for example be able to translate the English \emph{bank} into Dutch \emph{oever} ('river bank') when the words like \emph{boat} are mentioned, but into \emph{bank} (financial institution) when \emph{money} is mentioned.

@@

Assuming such a breakthrough will indeed come, the relevance for society is a big one: by improving the quality of machine translation, one basically improves (1) to what extent someone can get information from a document written in a human language he or she does not master, and (2) to what extent people without a common language can communicate. As for the first point, a lot of the roughly 7000 languages in the world also have a written form. This means that, unless you know all of these languages, part of all information available can only be reached by translation. This can of course be done a translator, but that's a time-consuming and labour-intensive job, which forces us to concentrate on translating only the more important documents to the more important languages. With good machine translation, we can easily make available \emph{all} information available to everyone. The second point is perhaps even more important: if for two people there isn't a language they both speak, they have very little ways to communicate. In practice this means whole communities of people are excluded from the rest of the world, simply because they do not know the languages of the rest of the world. In highly educated societies this is of course less of a problem, as people can easily learn a new language as long as they are willing to do some effort, but foreign language education is not available in large parts of the world. If it manages to get good enough, machine translation might save the day here: by building 'linguistic bridges' between the world's communities, it creates the possibility for everyone on the globe to communicate with everyone on the globe. This means one is no longer limited to the community he or she was born, which may increase one's chances for a happy and succesful life significantly.

@@

An idea is very different from something that is really useful or true. Many fields of science get stuck somewhere in this transition. In some fields scholars only write down their ideas in human language, not fully formulating how the idea should work, and thus never discovering there might be flaws in the idea somewhere. In other fields scholars can make the idea formal, but have no way to test the ideas in any other way than trusting on intuitions. In again other fields one can test it, for example by means of experiments, but it remains unclear how the results should be interpreted and to what extent they generalize to the whole population. All this is not a problem for the Colibri project. Claims are made (for example 'skipgrams improve machine translation'), formalized ('this is how they should work within in a machine translation system'), but also implemented and tested in a fully functional machine learning system. In other words, except maybe for the evaluation measures, there is little reason to doubt the results of this project.

@@

When describing the rule-based approaches to machine translation, I pointed out that for most approaches one needs a specific ruleset for each language pair. The Interlingua approach is an attempt to solve this problem by adding a language in between, so that only a ruleset for every language-Interlingua pair is needed. Statistical and memory-based approaches solve this problem in an even more elegant way: the only things needed are computational power, time, and a lot of data. As digital translated texts are available in large quantities for the more important languages of the world, adding more languages to Colibri should be relatively easy. But even less important languages can be added easily, as long as there is enough data to train the classifiers (but see the first two weaknesses). Obviously, the opposite is true for the rule-based approach.

@@

Memory-based learning takes the rule-creating process out of the hands of the authors. This can be a good thing, because it means translation systems can be made without any knowledge of the source or the target language, but it also means the authors loses control over what the system produces. That is, if the system makes errors, solving this problem is not as simple as changing how the system works. \\\indent
Instead, the task of the creator now is to provide enough bilingual texts for training - preferably as much as possible - as this improves the change of finding a matching n-gram. And importantly, to avoid errors, this data should be of high quality. One incorrect translation, if not found anywhere else in the corpus, can already end up in the translation results. For example, when translating \emph{Will Justin Bieber ever hit puberty} from English to Vietnamese with Google Translate, which is also based on bilingual corpora, the result is \emph{Justin Bieber se bao gio den tuoi day thi}, which means as much as 'Justin Bieber will never hit puberty'.

@@

Memory-based learning generated a translation on the basis of context. This works well for languages in which word order is largely fixed, like English: for these languages, if you change the position of a particular word in the sentence (and thus the context), you also change the meaning of that sentence. There are languages, however, with relatively free word order; Hungarian, Russian and Finish, to name a few. In these languages, changing the position and thus the context of a particular word or phrase doesn't (really) change the meaning of the sentence. In other words, in these languages the main source of the knowledge of memory-based learning isn't fully available, which may very well mean that memory-based translation may not be useful for these languages. To what extent this really is the case is completely unclear. \\\indent

>>>>

When one of Starfleet's vessels meets an alien ship in areas where no man has gone before, they can (almost) always communicate with these aliens (almost) instantly; they speak to them in English, and these aliens answer in English. This is possible because, according to the Star Trek lore, the protagonists make use of a so-called Universal Translator. This Universal Translator can (1) recognize speech in every language, (2) translate what it has 'heard' to any other language, (3) make its results audible with speech synthesis and (4) learn itself new languages when needed. Unfortunately, like many technologies in Star Trek, this is far from reality: there is lots of active research at the first three tasks in particular (and even some of the fourth, see \citet{biemann11}), but no software currently in existence can do this as perfect as the Universal Translator. \\\indent
In this paper, I will give a detailed overview of the research project Colibri, which basically is an attempt to improve the technology needed for the second task: automatically translating text from one language to another. More specifically, the project investigates the effects of various translation units; for example, whether making use of constructions with one or more gaps or context improves translation quality. Importantly, the translation is not done on the basis of explicit 'human' knowledge about language and grammar ('linguistic theory'), but are distilled from large amounts of text ('text corpora'). The Colibri project is carried out by Maarten van Gompel and supervised by Antal van den Bosch.
\\\indent
In section 2, I will give a short overview of what has been achieved in machine translation, and in section 3 the current project will be described. This order was chosen to show that Colibri builds on previous research directly. I give more information on the methodology in section 4, and describe the current and future station section 5. Section 6 elaborates on Colibri's relevance, both for science and society, and section 7 describes some strengths and weaknesses of the project. The paper is concluded in section 8.

@@

The horizontal arrows represent the approaches. I will go through them from bottom to top:

\begin{itemize}
\item \emph{The direct approach} does very little source language analysis, and could thus be viewed as the automated version of looking up every word in the source language in a dictionary and replacing it with the corresponding word in the target language. This often does not yield very good results: if I for example would have translated this paper from my mother tongue Dutch to English word for word, 'would we now not real a good result have'. It should be noted, however, that the direct approach mostly is more complicated than this; for example, in some cases morphological analysis is needed before it can be looked up in a dictionary, and more morphological analysis is needed for the translation word to fit in the sentence. Word reordering rules can also be included.
\item \emph{The syntactic approach} tries to discover the syntactic structure of the source sentence before it starts translating, using a parser. This structure is then translated to the target language, and filled with translated words.
\item \emph{The semantic approach} goes even one step further and also tries to grasp the meaning of a sentence, roughly. A sentence like \emph{Mummy kisses daddy} for example entails that mummy is the agent, and daddy the patient of the sentence. After having identified these semantic roles, the algorithm would then look how this meaning is expressed in the target language, find the approriate syntactic structure for that, and fill it with the translated words.
\item \emph{The interlingua approach}. All of the approaches so far have in common that they require a separate rule set for each language pair; no or very few rules included for translations from English to German can also be used for translations from English to French. This means that for every time a new language is added to the system, the amount of rules that have to be added grows exponentially. The Interlingua approach is an attempt to solve this problem, by pairing up every language with a formal representation of meaning, the Interlingua. This means a translation from English to German would entail a translation from English to the Interlingua (using the ruleset English $\rightarrow$ Interlingua) and then a translation from the Interlingua to German (using the ruleset Interlingua $\rightarrow$ German). When translating from English to French, the same two steps would be taken, and the same ruleset English $\rightarrow$ Interlingua can be used.
\end{itemize}

I should add that translation systems focussing on one approach are rare. As \citet{jm09} put it: 'Real systems tend to involve combinations of elements from these three architectures; thus, each is best thought of as a point in an algorithmic design space rather than as an actual algorithm.' The Systran system \citep{hs92,senellartea01}, for example combines the first three approaches.\\\indent
An advantage of the rule-based approach is that one has a lot of control over the output. In theory, if something has gone wrong, all one has to do is find out which rule or dictionary item caused the problem and improve it. In practice, however, it is almost impossible to have a dictionary and rule system good enough for more complicated sentences, and things like rule interactions and ambiguity make everything even more complicated. The next two approaches therefore replace hand-crafted rules with automatically generated ones.

@@

Unlike the rule-based approach, the statistical approach creates its own 'knowledge', using bilingual text corpora. This knowledge is probabilistic: instead of using fixed rules, statistical machine translation systems only know how likely a particular translation is on the basis of the data it has seen before. On the basis of these data, various translation hypotheses generated, from which the best one is chosen according to various statistical measures. I will explain these stages in more detail:

\begin{itemize}
\item \emph{Corpus alignment.} Only having a bilingual corpus is not enough: for such a corpus to be useful one also needs to know which parts of the text in language A correspond to which parts of the text in language B. This goes for both the sentence level (which sentence in language A corresponds to which sentence in language B), and the word level (which word in sentence X in language A corresponds to which word in sentence X in language B). Aligning words and sentences is complicated by the fact that sometimes language A uses more or less sentences/words than language B to express the same concept. After aligning words, it is possible to extract aligned phrases on the basis of word-alignment information. \citet{koehn03} show this provides better results than word-based translation.

\item \emph{Hypothesis generation.} We now have some sort of automatically generated dictionary for each word or phrase in language A, we know which word or phrase in language B is often used for it, and with what probability. In phrase-based machine translation, this dictionary is called the \emph{phrase translation table}. On the basis of this dictionary or phrase translation table, we can generate various possible translations. \\\indent

@@

\item \emph{The language model.} The language model predicts to what extent the generated sentence is expected for the target language; in other words, is this sentence a well-formed sentence or not? Because the unlimited number of ways words can be combined into sentences, it is unlikely the sentence under investigation will actually be found in a corpus, so the language model tries to do this prediction on the basis of how often smaller parts of the sentence (n-grams) occur in a corpus in sequence.

@@

The algorithm that performs the last two steps, hypothesis generation and evaluation, is called a \emph{decoder}. I should add that, in practice, not all possible hypotheses will be generated and evaluated, because that would take too much time and memory. Instead, the decoder's task could be viewed as search problem: we have an enormous pile of possible hypotheses of which one is this best, and we want to find that one hypothesis as quickly and efficiently as possible. To achieve this, only a small set of hypotheses is generated and evaluated immediately. New hypotheses are only generated on the basis of the outcomes, elimating search directions which are very unlikely to produce good results. In other words, in actual translation systems hypothesis generation and evaluation are mostly intertwined.

@@

For my overview of the example-based approach, I will focus on the memory-based version of the example-based approach in particular, because that is the version the Colibri project uses. Other versions of example-based machine translation might vary slightly. \\\indent
The memory-based approach is in many ways similar to the statistical approach - it could even easily be argued to be a part or an extension of it. For example, the memory based approach also generates translations on the basis of large bilingual corpora, and also works with a decoder to mold the various translation pieces found into a sentence. The main difference, however, is that it takes the context of the source language into account; the idea is that the system will find another translation for \emph{left} in the sentence \emph{The toilets are at your left hand} than for \emph{left} in the sentence \emph{Elvis has left the building} simply because they occur in different contexts. \\\indent
In the memory-based approach, this context-sensitivy is achieved with the help of machine learning: by showing a so-called classifier lots of examples of something, and telling it to which class it belongs, the classifier is able to predict the class of new examples. For example, we can give a classifier the external characteristics of 10.000 humans (length, type of cloths, length of hair, etc.), and also tell it whether these humans are male or female. If we then ask for the sex of a relatively short human, wearing a dress, with long hair, the classifier will probably be able to tell us this is a woman, because most of the examples with similar features also were women. It thus is frequency that plays a central role here.\\\indent
While this example only has two classes, it is in principle possible to use a very large number of classes. This is exactly what the memory-based approach to machine translation does: on the basis of a particular word and its context (for example, its direct neighbours), a classifier in machine translation returns as a class the most likely translation. Thus, instead of on human characteristics the classifiers are trained on words and their context from the source language, and instead of sex it uses n-grams from the target language as classes. \\\indent
There are two types of classification methods: \emph{eager learning} and \emph{lazy learning}. Whereas eager learning algorithms try to create a small, abstract, general model, filtering out infrequent cases, lazy learning algorithms take into account everything they encounter. According to \citet{dvdb05}, for natural language processing the lazy learning approach works best, because infrequent cases are an important part of the knowledge needed. \\\indent
Like I did with statistical machine translation, I will now give an overview of the memory-based machine translation process, adapted from \citet{vdbb09}. For this example, I will describe a translator that uses trigrams, but all kinds of n-grams could be used here, even n-grams of variable length. In fact, \citet{vangompel09}, \citet{vangompelea09} and \citet{vangompel11} showed that better results can be achieved using phrases of variable length; investigating the nature of multi-word units in translation therefore is one of the main goals of the Colibri project.

\begin{itemize}
\item \emph{Training.} Training is done only once and \emph{before} actual translation, and could therefore be seen as 'preparing' the translation software for its task. It consists of (1) aligning the bilingual corpus, using the word alignment algorithms described in the previous sections, (2) decomposing the complete corpus into trigrams, and (3) creating a translation model on the basis of this. The trigrams of the source language constitute the examples, the trigrams of the target language are the classes.

\item \emph{Local classification.} Classification means (1) decomposing the source text into trigrams and (2) classifying each trigram, based on the translation model created in the training phase. The output are the classes, i.e. the trigrams, the classifiers associates with each source trigram.

@@

Please note again the similarities with the statistical approach: like in the \emph{hypothesis generation} phase, (parts of) hypotheses are generated during \emph{local classification}, and like the \emph{hypothesis evaluation phase}, hypotheses are evaluated during \emph{global search}.

@@

The gap here is filled with 'his', but the construction 'from \_\_ point of view' can also be filled with many other words ('my', 'the', 'another', 'Obama's', etc.). This shows the appeal of skipgrams over n-grams for machine translation: whereas n-grams would only have been able to capture the construction 'point of view', because the word directly before 'point of view' is variable, skipgrams are also able to recognize that the word 'for' is part of the construction. Frequent examples of the skipgrams (I expect the variant with 'my' to be very frequent, for instance) of course \emph{could} be recognized by the n-gram approach.\\\indent
Of course, not every possible group of consecutive words is a construction; what makes some groups special? The exact nature of linguistic constructions has been the subject of many linguistic publications and even an entire group of linguistic frameworks (construction grammar, see \citet{goldberg95} for a well-known example). For this project, however, it suffices to say that constructions emerge because some combinations of words are more frequent than others. As we have seen in the previous section, taking into account frequency in the training material is how classifiers are able to do their job.

@@

\subsubsection{Identifying and extracting constructions in a language}
When using large corpora, the process of discovering simple n-grams might already take quite a lot of time and memory; adding variability in length and one or multiple gaps to the n-grams makes the process even more - in fact, many times more - complex. To reduce the processing load and resource consumption, two techniques are used:

\begin{itemize}
\item Optimization of how the various the n-grams and skipgrams are saved in memory. By reserving the labels that do not take a lot of memory for the most frequent grams, a lot of memory can be saved.
\item For trigrams and larger n-grams, taking into account the frequencies of the parts. If these parts do not exist or are very infrequent, there is no use in looking for the whole - if even the parts do not exist, the whole will not exist for sure and doesnt need to be investigated any further.
\end{itemize}

\subsubsection{Identifying aligned constructions between language-pairs}
Once the constructions in both parts of a bilingual corpus are identified, these constructions need to be linked. Colibri aligns constructions by aligning the individual words first and constructions later, and uses a separate algorithm to align skipgrams.\\\indent
It should be noted that this approach is the result of various investigations. For example, instead of the classical approach of aligning constructions by aligning the individual words first, as described by \citep{koehn03}, it was attempted to try to align the constructions directly. To achieve this, one can for example simply measure how often the constructions found in both languages occur together (that is, in both the source and the target language). Because at this stage the decoder was not yet ready (see subsection 4), the decoder from translation system MOSES \citep{koehnea07} was used to evaluate the results. These results showed that the approach using constructions directly for alignment could not provide results as good as the approach mediated through word alignments; therefore, it was decided to go back to the classical method. It should be noted that, because the MOSES decoder was used, the influence of skipgrams on alignment could not be measured. However, because they increase complexity, they are not expected to improve the alignments.\\\indent
When support for skipgrams was added to the implementation of the classical approach, it turned out that including skipgrams in the alignment process had negative effects on the results. To fix this, an algorithm specialized for aligning skipgrams was developed, which indeed was able to find good translations.

\subsubsection{Memory-based classification using construction experts}
As may have become clear from the previous sections, the Colibri project investigates the use of machine learning techniques to get from the source language construction to the target language constructions. More specifically, Colibri uses specialized classifiers, trained for only one construction, inspired by the effectivity of this approach in word sense ambiguation \citep{vangompel10}. The results of this approach are compared to the results of a combined classifier approach and statistical machine translation.

\subsubsection{Reassembling constructions into a coherent target sentence}
A decoder was developed from scratch, following state-of-the-art techniques as described by \citet{koehn03}. Support for skipgrams and machine learning, of course not available in standard decoders, was added.

@@

Because of Colibri's practical nature, its methodology forms the core of the project and has been explained in detail in the previous sections, making this section largely superfluous. The only thing I would like to give some attention is that Colibri's time planning is similar to that of other research projects:

\begin{enumerate}
\item reading the previous literature
\item building a theoretical model
\item testing the model, for example with experiments
\item refining the model on the basis of the results
\end{enumerate}

However, what is actually done during these stages is radically different from many other fields of science. For example, whereas stage 2 in many research projects will result in a paragraph of text, a table, a flowchart, a formula, or something similar, Colibri's stage 2 actually results in a fully functional translation system. And whereas stage 3 in many fields of science entails designing some kind of experiment that will hopefully provide evidence for the model developed during stage 2 (and then finding people to participate in that experiment), Colibri's stage 3 consists of writing a script to evaluate the results of the translation system, using various evaluation measures.

@@

At the moment, the functional machine translation system described in section \ref{builder} is mostly finished. It is mainly inspired upon the open source translation system MOSES, and indeed manages to produce similar results. However, as was explained in section \ref{Description}, Colibri is an attempt to improve the quality of machine translation by adding skipgrams and memory-based learning. Unfortunately it has turned out that both do not add much to the translation quality: memory-based learning only improves the results a little bit, and the skipgrams are mostly ignored by the decoder. Apparently, translation hypotheses without skipgrams are more likely to be good translations in most of the cases. \\\indent
This means Colibri's goal, improving translation quality, has not yet been achieved. It is possible that larger improvements in translation quality \emph{can} be achieved by changing things like corpora or even languages used, but if this is not the case, new possibilities for improvements have to be investigated. One such possibility is to include a system that finds the correct translations of words on the basis of particular keywords in a sentences. This system should for example be able to translate the English \emph{bank} into Dutch \emph{oever} ('river bank') when the words like \emph{boat} are mentioned, but into \emph{bank} (financial institution) when \emph{money} is mentioned.

@@

Assuming such a breakthrough will indeed come, the relevance for society is a big one: by improving the quality of machine translation, one basically improves (1) to what extent someone can get information from a document written in a human language he or she does not master, and (2) to what extent people without a common language can communicate. As for the first point, a lot of the roughly 7000 languages in the world also have a written form. This means that, unless you know all of these languages, part of all information available can only be reached by translation. This can of course be done a translator, but that is a time-consuming and labour-intensive job, which forces us to concentrate on translating only the more important documents to the more important languages. With good machine translation, we can easily make available \emph{all} information available to everyone. The second point is perhaps even more important: if for two people there is not a language they both speak, they have very few ways to communicate. In practice this means whole communities of people are excluded from the rest of the world, simply because they do not know the languages of the rest of the world. In highly educated societies this is of course less of a problem, as people can easily learn a new language as long as they are willing to do some effort, but foreign language education is not available in large parts of the world. If it manages to get good enough, machine translation might save the day here: by building 'linguistic bridges' between the world's communities, it creates the possibility for everyone on the globe to communicate with everyone on the globe. This means one is no longer limited to the community he or she was born, which may increase one's chances for a happy and succesful life significantly.

@@

An idea is very different from something that is really useful or true. Many fields of science get stuck somewhere in this transition. In some fields scholars only write down their ideas in human language, not fully formulating how the idea should work, and thus never discovering there might be flaws in the idea somewhere. In other fields scholars can make the idea formal, but have no way to test the ideas in any other way than trusting on intuitions. In again other fields one can test it, for example by means of experiments, but it remains unclear how the results should be interpreted and to what extent they generalize to the whole population. All this is not a problem for the Colibri project. Claims are made (for example 'skipgrams improve machine translation'), formalized ('this is how they should work within in a machine translation system'), but also implemented and tested in a fully functional machine learning system. In other words, except maybe for the evaluation measures, there is little reason to doubt the results of this project, because of its highly emperical nature.

@@

When describing the rule-based approaches to machine translation, I pointed out that for most approaches one needs a specific ruleset for each language pair. The Interlingua approach is an attempt to solve this problem by adding a language in between, so that only a ruleset for every language $\rightarrow$ Interlingua and Interlingua $\rightarrow$ language pair is needed. Statistical and memory-based approaches solve this problem in an even more elegant way: the only things needed are computational power, time, and a lot of data. As digital translated texts are available in large quantities for the more important languages of the world, adding more languages to Colibri should be relatively easy. But even less important languages can be added easily, as long as there is enough data to train the translation model (but see the first two weaknesses). Obviously, the opposite is true for the rule-based approach.

@@

Statistical and Memory-based machine translation takes the rule-creating process out of the hands of the authors. This can be a good thing, because it means translation systems can be made without any knowledge of the source or the target language, but it also means the authors loses control over what the system produces. That is, if the system makes errors, solving this problem is not as simple as changing how the system works. \\\indent
Instead, the task of the creator now is to provide enough bilingual texts for training - preferably as much as possible - as this improves the change of finding a matching n-gram. These data are simply not available for all languages in the world. And not only should there be enough bilingual texts, they should also be of high quality. One incorrect translation, if not found anywhere else in the corpus, can already end up in the translation results. For example, when translating \emph{Will Justin Bieber ever hit puberty} from English to Vietnamese with Google Translate, which is also based on bilingual corpora, the result is \emph{Justin Bieber se bao gio den tuoi day thi}, which means as much as 'Justin Bieber will never hit puberty'.

@@

Memory-based learning generates a translation on the basis of source-context. This works well for languages in which word order is largely fixed, like English: for these languages, if you change the position of a particular word in the sentence (and thus the context), you also change the meaning of that sentence. There are languages, however, with relatively free word order; Hungarian, Russian and Finnish, to name a few. In these languages, changing the position and thus the context of a particular word or phrase does not (really) change the meaning of the sentence. In other words, in these languages the main source of the knowledge of memory-based learning is not fully available, which may very well mean that memory-based translation may not be useful for these languages. On the other hand, in many languages with free word order there is something like a canonical word order (so, although the order is free, it is not random), and almost all languages with free word order use case instead. Case could be really useful to memory-based learning, because having a word in a particular case in the context might be a very good clue which possible translation is the best. To what extent free word order is a problem or case is helpful is completely unclear.
\\\indent
